# Synthetic timeseries data cases

```{r}
library(tidyverse)
library(patchwork)
source('functions/adj_gam_sim.R')
```

# Simulate some data

We will simulate data for this example because we simply haven't the time to be dealing with the horrors of real world datasets. We will be creating a synthetic timeseries of streamflow data which is modelled with additive contributions from non linear transformations of common covariates.

First lets define the functions we will be using to generate the data.

```{r}
set.seed(2319)
source('functions/adj_gam_sim.R')
gt_data <- generate_gtdata()

# plot the functions we are trying to fit
ggplot() +
    stat_function(fun = f0, aes(color = 'f0')) +
    # stat_function(fun = f1, aes(color = 'f1')) +
    stat_function(fun = f2, aes(color = 'f2')) +
    stat_function(fun = f3, aes(color = 'f3')) +
    stat_function(fun = f4, aes(color = 'f4')) +
    geom_line() +
    theme_minimal()
```

## Load covariates

We are using SOI and rainfall data from BOM to generate a streamflow timeseries.

BOM SOI can be downloaded from (just save in `raw_data/soi.txt`):
http://www.bom.gov.au/clim_data/IDCKGSH000/soi_monthly.txt

BOM rainfall data can be downloaded from climate data online (just save the data csv file as `raw_data/66006.csv`):
http://www.bom.gov.au/climate/data/ and search for 66006 in "3. Get the data"

TODO: Work out the data conditions of use to see if I can just provide the csv synthetic data.

```{r}
# load SOI
soi_data <- read_csv(
    paste("raw_data","soi.txt", sep = .Platform$file.sep),
    col_names = c("yyyymm", "SOI")
)
soi_data <- soi_data %>% mutate(
    month = ym(yyyymm)
)
```

```{r}
# Load
rainfall_data <- read_csv(
    paste("raw_data", "66006.csv", sep = .Platform$file.sep)
)

rainfall_data <- rainfall_data %>% 
    select(c("Year", "Month", "Day", "Rainfall amount (millimetres)")) %>%
    rename(y = "Year", m = "Month", d = "Day", rain = "Rainfall amount (millimetres)") %>%
    mutate(rain = ifelse(is.na(rain), 0, rain)) %>%
    mutate(
        date = ymd(paste(y, m, d, sep = "-")),
        month = as.Date(floor_date(date, "month"))
    ) %>% summarise(
        rain = sum(rain),
        .by = month
    ) %>% left_join(
        soi_data %>% select(month, SOI),
        by = "month"
    ) %>% mutate(
        year = as.factor(year(month)),
        moy = lubridate::month(month),
        time = (lubridate::year(month) - lubridate::year(min(month)))*12 + lubridate::month(month)
    )
```

## Modify the covariates data

We are going to be devious and modify the data for maximum effect in the example - so we pick a period so that SOI and time are vaguely collinear.

```{r}
# and chop to some shorter more ambuiguous period
rainfall_data <- rainfall_data %>% filter(month >= ym("198301"), month <= ym("202312"))
# scale the variables so that we can use our nice f0-f4 functions
rainfall_data <- rainfall_data %>% mutate(
    time = time - min(time)
) %>% mutate(
    time = time/max(time), # time into 0-1,
    moy = moy/max(moy), # month into 0-1
    rain = (rain - min(rain))/(max(rain) - min(rain)), # rain into 0-1
    SOI = (SOI - min(SOI))/(max(SOI) - min(SOI)) # SOI into 0-1
)

head(rainfall_data)
```

## Plot the covariates data

```{r}
# Annual summary plot
ggplot(rainfall_data %>% summarise(month = first(month), rain = sum(rain), .by = year), aes(x = month, y = rain)) +
    geom_bar(stat = "identity", width=250) +
    labs(
        x = "Year",
        y = "Rainfall (mm)"
    )
ggplot(rainfall_data %>% summarise(month = first(month), SOI = mean(SOI), .by = year), aes(x = month, y = SOI)) +
    geom_line(stat = "identity", width=250) +
    labs(
        x = "Year",
        y = "SOI"
    )
```

# Generate the synthetic timeseries
Here we take our covariates and use our target smooths to create some vaguely realistic synthetic dataset.

```{r}
set.seed(2319)
scale <- 0.25
streamflow_data <- rainfall_data %>% tsSim(scale=scale)

wrap_plots(
    ggplot(streamflow_data, aes(x = month, y = flow)) +
        geom_line() +
        labs(
            x = "Year",
            y = "Streamflow (ML)"
        ),
    ggplot(streamflow_data, aes(x = month, y = flow)) +
        geom_line() +
        #log scale
        scale_y_log10() +
        labs(
            x = "Year",
            y = "Log Streamflow (ML)"
        ),
    ggplot(streamflow_data, aes(x = month, y = rain)) +
        geom_bar(stat = "identity") +
        labs(
            x = "Year",
            y = "Rainfall (mm)"
        ),
    ncol = 1
)

# plot correlation to rainfall, soi
wrap_plots(
    ggplot(streamflow_data, aes(x = rain, y = flow)) +
        geom_point() +
        labs(
            x = "Rainfall (mm)",
            y = "Streamflow (ML)"
        ),
    ggplot(streamflow_data, aes(x = SOI, y = flow)) +
        geom_point() +
        labs(
            x = "SOI",
            y = "Streamflow (ML)"
        ),
    ncol = 2
)
```

## Save the data

```{r}
fpath <- paste("synthetic_data", paste0("synthetic_streamflow_data.csv"), sep = .Platform$file.sep)
# ensure it exists
dir.create("synthetic_data", showWarnings = FALSE)
write_csv(streamflow_data, fpath)
```